# forecastxgb-r-package
An R package for time series models and forecasts with xgboost compatible with {forecast} S3 classes

Only on GitHub.  Very early days, incomplete.  Planned addition is support for `xreg`.

This implementation uses as features lagged values of the target variable, linear time, and dummy variables for seasons.

```{r eval = FALSE}
devtools::install_github("ellisp/forecastxgb-r-package/pkg")
```

## Usage
Seems to overfit rather severely, judging from the in-sample accuracy.  This is despite there being a cross-validation step:
```{r}
library(forecastxgb)
model <- xgbts(AirPassengers)
fc <- forecast(model, h = 12)
accuracy(fc)
plot(fc)

```


## Tourism data example
```{r}
library(Tcomp)

thedata <- tourism[[1]]

x <-thedata$x
h <- thedata$h

model <- xgbts(x)
xgbts_importance(model)


fc <- forecast(model, h = h)
plot(fc, bty = "l")
lines(thedata$xx, col = "red")
legend("topleft", legend = c("xgb forecast", "actual"), lty = 1, col = c("blue", "red"), bty ="n")
accuracy(fc, thedata$xx)
```

## Non-seasonal data

```{r}
obj <- xgbts(Nile)
xgbts_importance(obj)
fc <- forecast(obj, 30)
plot(fc, bty = "l")
```

